---
title: "Horse Profiles Data Visualizations"
output: 
  html_document:
    theme: simplex
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
    collapse: TRUE
date: "`r Sys.Date()`"
---

The goal of this document is to visualize the horse profile data to get a sense of what our sample of horses looks like. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center", out.width = '100%')
library(tidyverse)
library(viridis)
library(GGally)
library(patchwork)
library(pointblank)
library(reactable)
library(reactablefmtr)
library(ggbeeswarm)
```

----

# Loading in the Data

Each row in the data set represents a unique horse. The columns represent specific variables of interest about that horse. We have 1679 horses and 36 variables of interest.

```{r}
horse_profiles <- read_csv("./outputted-csvs/horse_profiles.csv")
```


```{r}
horse_profiles %>% 
  group_by(if_injury_reported, fatal) %>% 
  count() %>% 
  knitr::kable()
```
----

# Introductory Relationships

In this first section, we explore some initial univariate relationships in the horse profiles data set.

### Number of Times a Horse Raced

```{r}
horse_profiles %>% 
  ggplot(aes(x = n_races_tracked_2019))+
  geom_histogram(binwidth = 1)+
  scale_y_continuous(breaks = seq(0,1500,100))+
  labs(x = "Number of Races", y = "Count", title = "All Races")+
  theme_minimal()
```

### Horses' Average Race Distance (in Meters)

```{r}
horse_profiles %>% 
  ggplot(aes(x = avg_race_dist_meters))+
  geom_histogram()+
  scale_y_continuous(breaks = seq(0,2000, 200))+
  labs(x = "Average Race Distance per Horse (in Meters)", y = "Count", title = "All Races")+
  theme_minimal()
```

### Average Total Strain per Horse

```{r}
horse_profiles %>% 
  ggplot(aes(x = avg_total_race_strain))+
  geom_histogram()+
  scale_y_continuous(breaks = seq(0,2000, 50))+
  scale_x_continuous(breaks = seq(-100,200,20))+
  labs(x = "Average Total Race Strain per Horse", y = "Count", title = "Mile Races")+
  theme_minimal()
```

### Frequency of Injuries

```{r}
horse_profiles %>% 
  ggplot(aes(x = if_injury_reported))+
  geom_bar()+
  scale_y_continuous(breaks = seq(0,4500,500))+
  labs(y = "Count", x = "The Horse had a Reported Injury", title = "All Races")+
  theme_minimal()
```

### Number of Races by Injured vs. Not Injured Horses


```{r}
horse_profiles %>% 
  mutate(longform_injured = if_else(if_injury_reported == TRUE, "Injured", "Not Injured")) %>% 
  ggplot(aes(y = longform_injured, x = n_races_tracked_2019))+
  geom_violin(aes(fill = longform_injured), alpha = .1)+
  geom_boxplot(aes(color = longform_injured),width = .3)+
  labs(y = "", x = "Number of Races", title = "Distribution of Races Count for\nInjured versus Non-Injured Horses\n")+
  scale_x_continuous(breaks = seq(0,25,2))+
  theme_minimal()+
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16), title = element_text(face = "bold", size = 12),
        axis.text = element_text(color = "black", size = 12))
```


### Horse Fatalities

```{r}
horse_profiles %>% 
  filter(if_injury_reported == TRUE) %>% 
  ggplot(aes(x = fatal))+
  geom_bar()+
  theme_minimal()+
  labs(x = "Fatal?", y = "Count", title = "All Races")+
  theme(legend.position = "none")
```


### Training, Racing, or Other...Where Did Injury Occur?


```{r}
horse_profiles %>% 
  filter(if_injury_reported == TRUE) %>% 
  ggplot()+
  geom_bar(aes(x = injury_location))+
  labs(x = "Injury Location", y = "Count", title = "Number of Injuries at Each Racing Track")+
  theme_minimal()+
    theme(legend.title = element_blank(), legend.position = "bottom", 
        legend.justification = "right",
        legend.text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16), title = element_text(face =
        "bold", size = 12), axis.text = element_text(color = "black", size = 12))
```

### Race Track at which Injury Occured

```{r}
horse_profiles %>% 
  filter(if_injury_reported == TRUE) %>% 
  ggplot()+
  geom_bar(aes(x = injury_track))+
  scale_y_continuous(breaks = seq(0,60,10))+
  labs(x = "Racing Track", y = "Count", title = "Number of Injuries at Each Racing Track")+
  theme_minimal()+
    theme(legend.title = element_blank(), legend.position = "bottom", 
        legend.justification = "right",
        legend.text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16), title = element_text(face =
        "bold", size = 12), axis.text = element_text(color = "black", size = 12))
```

### Days from Injury to Next Race

```{r}
horse_profiles %>% 
  filter(if_injury_reported == TRUE) %>% 
  ggplot()+
  geom_histogram(aes(x = days_injury_to_race), bins = 20)+
  labs(x = "Days from Injury to Next Race", y = "Count", title = "All Races")+
  theme_minimal()
```


### Changes in Number of Races per Horse between 2019 to 2021

```{r}
horse_profiles %>% 
  select(horse_id:n_races_2021, if_injury_reported, -n_races_tracked_2019) %>%
  pivot_longer(n_races_2019:n_races_2021, names_to = "year", values_to = "value") %>%
  drop_na(value) %>% 
  ggplot(aes(y = year, x = value))+
  geom_violin(aes(fill = year), alpha = .3)+
  geom_boxplot(width = .2, aes(color = year))+
  theme_minimal()+
  theme(legend.position = "none")
  
```

```{r}
horse_profiles %>% 
  select(horse_id:n_races_2021, if_injury_reported, -n_races_tracked_2019) %>%
  mutate(n_diff_2019_2020 = n_races_2020 - n_races_2019) %>% 
  mutate(n_diff_logical = if_else(n_diff_2019_2020 < 0, TRUE, FALSE)) %>% 
  group_by(n_diff_logical) %>% 
  count() %>% 
  knitr::kable()
```



```{r}

horse_id_bottom <- horse_profiles %>% 
  select(horse_id:n_races_2021, if_injury_reported, -n_races_tracked_2019) %>%
  mutate(n_diff_2019_2020 = n_races_2020 - n_races_2019) %>% 
  filter(n_diff_2019_2020 <  quantile(n_diff_2019_2020, probs = 0.05, na.rm = TRUE))



p1 <- horse_profiles %>% 
  select(horse_id:n_races_2021, if_injury_reported, -n_races_tracked_2019) %>%
  pivot_longer(n_races_2019:n_races_2021, names_to = "year", values_to = "value") %>%
  drop_na(value) %>% 
  mutate(year = case_when(
    str_detect(year, "2019") ~ 2019,
    str_detect(year, "2020") ~ 2020,
    str_detect(year, "2021") ~ 2021,
    TRUE ~ NA
  )) %>% 
  filter(horse_id %in% horse_id_bottom$horse_id) %>% 
  ggplot(aes(x = year, y = value))+
  geom_line(aes(group = as_factor(horse_id), color = as.factor(horse_id)), linetype = "dashed")+
  geom_point(aes(group = as_factor(horse_id), color = as.factor(horse_id)), size = 2.5)+
  scale_x_continuous(breaks = seq(2019,2021,1))+
  scale_y_continuous(breaks = seq(0,30,2))+
  labs(x = "Year", y = "Number of NYRA Races Started", title = "Bottom 5% of Difference in\nNumber of Races in 2019 to 2020")+
  guides(color = guide_legend(title = "Horse ID"))+
  theme_minimal()+
  theme(legend.position = "none")
```

```{r}
horse_id_top <- horse_profiles %>% 
  select(horse_id:n_races_2021, if_injury_reported, -n_races_tracked_2019) %>%
  mutate(n_diff_2019_2020 = n_races_2020 - n_races_2019) %>% 
  filter(n_diff_2019_2020 >  quantile(n_diff_2019_2020, probs = 0.95, na.rm = TRUE))



p2 <- horse_profiles %>% 
  select(horse_id:n_races_2021, if_injury_reported, -n_races_tracked_2019) %>%
  pivot_longer(n_races_2019:n_races_2021, names_to = "year", values_to = "value") %>%
  drop_na(value) %>% 
  mutate(year = case_when(
    str_detect(year, "2019") ~ 2019,
    str_detect(year, "2020") ~ 2020,
    str_detect(year, "2021") ~ 2021,
    TRUE ~ NA
  )) %>% 
  filter(horse_id %in% horse_id_top$horse_id) %>% 
  ggplot(aes(x = year, y = value))+
  geom_line(aes(group = as_factor(horse_id), color = as.factor(horse_id)), linetype = "dashed")+
  geom_point(aes(group = as_factor(horse_id), color = as.factor(horse_id)), size = 2.5)+
  scale_x_continuous(breaks = seq(2019,2021,1))+
  scale_y_continuous(breaks = seq(0,20,2))+
  labs(x = "Year", y = "Number of NYRA Races Started", title = "Top 5% of Difference in\nNumber of Races in 2019 to 2020")+
  guides(color = guide_legend(title = "Horse ID"))+
  theme_minimal()+
  theme(legend.position = "none")
```

```{r}
library(patchwork)

p1 + p2
```



```{r}
# I think we want to see if there is a significant drop off in race count from year to year

horse_profiles %>% 
  select(horse_id:n_races_2021, if_injury_reported, -n_races_tracked_2019) %>%
  mutate(n_diff_2019_2020 = n_races_2020 - n_races_2019,
         n_diff_2021_2020 = n_races_2021 - n_races_2020)
  

```



----

# Injury Relationships

```{r}
horse_profiles %>% 
  filter(if_injury_reported == TRUE) %>% 
  ggplot(aes(x = days_injury_to_race, y = injury_track))+
  geom_violin(aes(fill = injury_track), alpha = .1, width = 1.5)+
  geom_boxplot(aes(color = injury_track),width = .1)+
  labs(x = "Days from Injury to Next Race", y = "Count", title = "All Races")+
  theme_minimal()+
  theme(legend.position = "none")
```

```{r}
horse_profiles %>% 
  pivot_longer(cols = ends_with("between_races"), names_to = "summary_stat", values_to = "n_days") %>% 
  ggplot(aes(x = n_days, y = if_injury_reported))+
    geom_violin(aes(fill = if_injury_reported), alpha = .1)+
  geom_boxplot(aes(color = if_injury_reported),width = .2)+
  labs(x = "Number of Days between Races", y = "Has Been Injured?", title = "All Races")+
  theme_minimal()
```

### DNFs versus Number of Injuries 

```{r eval = FALSE}
counts <- horse_profiles %>% 
  group_by(horse_id) %>%   
  summarize(n_injuries = sum(n_injury_type)) %>% 
  group_by(n_injuries) %>% 
  count()

horse_profiles %>% 
  group_by(horse_id) %>% 
  summarize(n_injuries = sum(n_injury_type), n_dnf = sum(n_dnf)) %>% 
  ggplot(aes(x = n_injuries, fill = as_factor(n_dnf)))+
  geom_bar(position = "fill")+
  geom_text(data = counts, aes(x = n_injuries, y = c(.9,.9,.9), label = paste0("Count: ",n)), inherit.aes = FALSE, color = "white")+
  labs(y = "Proportion")+
  guides(fill = guide_legend(title = "Number of DNFs"))+
  labs(x = "Number of Injuries per Horse", title = "All Races")+
  theme_minimal()
```

```{r}
set.seed(47)

cols = c("Days between Last Race and Injury" = "red")

horse_profiles %>%
  mutate(days_between = as.numeric(injury_date - last_race_pre_injury)) %>% 
    filter(if_injury_reported == TRUE & !is.na(days_between), !is.na(median_days_between_races), n_races_tracked_2019 > 3) %>% 
   slice_sample(n = 25) %>% 
  pivot_longer(cols = ends_with("between_races"), names_to = "summary_stat", values_to = "n_days") %>% 
  ggplot(aes(y = as.factor(horse_id), x = n_days))+
  geom_line()+
  geom_boxplot()+
  geom_point(aes(y = as.factor(horse_id), x = days_between, color = "Days between Last Race and Injury"), size = 2, inherit.aes = FALSE)+
  labs(x = "Number of Days between Races", y = "Horse ID", title = "Comparison between a Horse's Distribution of Days Between Races\nand the Number of Days Between their Last Race and Injury")+
  scale_color_manual(values = cols)+
  theme_minimal()+
  theme(legend.title = element_blank(), legend.position = "bottom", 
        legend.justification = "right",
        legend.text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16), title = element_text(face =
        "bold", size = 12), axis.text = element_text(color = "black", size = 12))
```

### Comparison between Race Distance (in Meters) of Average versus Injury Race 

```{r}
set.seed(47)

horse_profiles %>%
  filter(if_injury_reported == TRUE & !is.na(race_length_injury_meters) & n_races_tracked_2019 > 1) %>% 
  slice_sample(n = 45) %>% 
  mutate(diff_meters = avg_race_dist_meters - race_length_injury_meters) %>% 
  mutate(diff_meters_logical = if_else(diff_meters < 0, "Negative", "Positive")) %>% 
  ggplot(aes(y = as.factor(horse_id), x = diff_meters))+
  geom_bar(aes(fill = fatal, color = fatal),stat = "identity")+
  theme_minimal()+
  labs(y = "Horse ID", x = "Race Distance Difference (in Meters)", title = "The Difference between a Horse's Average Race Distance\nand their Race Distance where They got Injured")+
  guides(fill = guide_legend(title = "Fatal Injury?"), color = guide_legend(title = "Fatal Injury?"))+
  scale_x_continuous(breaks = seq(-400,300,50))+
  theme(legend.text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16), title = element_text(face =
        "bold", size = 12), axis.text = element_text(color = "black", size = 12))
```


```{r}
horse_profiles %>% 
  select(-horse_name, -horse_id, -injury_track_location_clean, - injury_details_reported, -third_quartile_days_between_races, -first_quartile_days_between_races, -avg_total_race_strain, -n_dnf, -n_injury_type) %>% 
  GGally::ggcorr(label = TRUE,
                 size = 2.95,
                 hjust = .85,
                 color = "black",
                 face = "bold",
                 legend.size = 10)+
  labs(title = "Correlations between Horse Profiles' Variables")+
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16), title = element_text(face =
        "bold", size = 12), axis.text = element_text(color = "black", size = 12))
```

```{r}
p1 <- horse_profiles %>% 
  ggplot(aes(y = days_off_pre_injury, x = avg_finishing_place))+
  geom_point()+
  geom_smooth()+
  scale_x_continuous(breaks = seq(0,10,1), limits = c(0.8,8.5))+
  theme_minimal()

p2 <- horse_profiles %>%
  ggplot(aes(y = days_injury_to_race, x = avg_finishing_place))+
  geom_point()+
  geom_smooth()+
  scale_x_continuous(breaks = seq(0,10,1), limits = c(0.8,8.5))+
  theme_minimal()

library(patchwork)
p1 + p2
```

```{r}
horse_profiles %>% 
  filter(if_injury_reported == TRUE) %>% 
  ggplot(aes(x = injury_track))+
  geom_bar(aes(fill = injury_track_location_clean), color = "black",position = "fill")+
  scale_y_continuous(breaks = seq(0,1,.1))+
  labs(x = "Race Track", y = "Proportion", title = "Comparison between the Injury Location \nProportions for Different Race Tracks")+
  guides(fill = guide_legend("Injury Location"))+
  theme_minimal()+
    theme(legend.text = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16), title = element_text(face =
        "bold", size = 12), axis.text = element_text(color = "black", size = 12))
```


```{r}
profiles_grouped_injuries_all <- horse_profiles %>% 
  mutate(injury_details_reported = str_to_lower(injury_details_reported)) %>% 
  mutate(injury_type = case_when(
    str_detect(injury_details_reported, "fracture|sesamoid") ~ "Fracture",
    str_detect(injury_details_reported, "laceration|bleed") ~ "Laceration",
    str_detect(injury_details_reported, "fell|stumbled|collapsed") ~ "Fall",
    str_detect(injury_details_reported, "tendon") ~ "Tendon",
    str_detect(injury_details_reported, "clipped|hit|^kick") ~ "Hit",
    str_detect(injury_details_reported, "laminitis") ~ "Muscle",
    str_detect(injury_details_reported, "suspensory")  ~ "Ligament",
    str_detect(injury_details_reported, "elbow|fetlock") ~ "Joint",
    str_detect(injury_details_reported, "colic|colon|pneumonia") ~ "Organ",
    str_detect(injury_details_reported, "head") ~ "Head",
    str_detect(injury_details_reported, "^vanned off|vanned off$|unknown|^pulled up|precaution|^injury$| no injury reported|lf$|broke down|flipped in gate|leg$|^injured$") ~ "Unspecified",
    TRUE ~ NA
  ))

## might be good to review the groupings above because I am not totally confident on them

profiles_grouped_injuries_all %>% 
  filter(if_injury_reported == TRUE) %>% 
  group_by(injury_type, fatal) %>% 
  count() %>% 
  ggplot()+
  geom_bar(aes(x = reorder(injury_type, -n), y = n, fill = fatal), stat = "identity")+
  scale_y_continuous(breaks = seq(0,80,10))+
  labs(x = "Injury Type", y = "Count", title = "All Races")+
  guides(fill = guide_legend(title = "Fatal?"))+
  scale_fill_hue(l = 50)+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5))
```